\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows, positioning}

\geometry{margin=1in}

\title{Thesis Progress Report: Proposed Research Architecture for Scientific Article Similarity Checking}
\author{Shinthiya Nowsain, Promi \\ Supervisor: Prof. Dr. Diana Kalibatienė}
\date{March 11, 2026}

\begin{document}

\maketitle

\section{Overview}
This report outlines the technical progress and the proposed architectural design for the second phase of the Master's Thesis. Following the approved methodology in Part 1, the implementation focuses on a \textbf{Hybrid Similarity Model} combined with a \textbf{Sequence Alignment Layer} to detect "Intelligent Plagiarism" in scientific literature.

\section{Proposed Research Architecture}
The system is designed as a three-tier pipeline to move beyond simple string matching and address the "semantic gap" in current detection tools.

\subsection{Visual Architecture (BPMN-style)}
\begin{center}
\begin{tikzpicture}[node distance=2cm]
    \tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm, minimum height=1cm,text centered, draw=black, fill=red!30]
    \tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm, text centered, draw=black, fill=orange!30]
    \tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm, text centered, draw=black, fill=green!30]
    \tikzstyle{arrow} = [thick,->,>=stealth]

    \node (in) [startstop] {Document Input};
    \node (prep) [process, below of=in] {Sliding Window Preprocessing};
    \node (hybrid) [process, below of=prep, yshift=-0.5cm] {\textbf{Hybrid Feature Extraction}};
    \node (sem) [process, left of=hybrid, xshift=-3cm] {SBERT (Semantic)};
    \node (str) [process, right of=hybrid, xshift=3cm] {POS/Structural (Syntactic)};
    \node (align) [process, below of=hybrid, yshift=-1cm] {\textbf{Smith-Waterman Alignment Layer}};
    \node (out) [startstop, below of=align] {SOTA Detection Result};

    \draw [arrow] (in) -- (prep);
    \draw [arrow] (prep) -- (hybrid);
    \draw [arrow] (hybrid) -- (sem);
    \draw [arrow] (hybrid) -- (str);
    \draw [arrow] (sem) |- (align);
    \draw [arrow] (str) |- (align);
    \draw [arrow] (align) -- (out);
\end{tikzpicture}
\end{center}

\subsection{Key Technical Components}
\begin{enumerate}
    \item \textbf{The Hybrid Model:} Integrates deep learning embeddings (\textbf{SBERT}) for conceptual meaning with handcrafted \textbf{Structural Features} (Noun/Verb ratios, sentence morphology). This fulfills Task 2 of the thesis.
    \item \textbf{Sequence Alignment Algorithm:} Implements the \textbf{Smith-Waterman} algorithm. Unlike simple cosine similarity, this algorithm treats the document as a sequence, identifying the longest continuous chain of similar ideas even if the paragraphs are reordered or "gaps" are inserted (patchwriting).
\end{enumerate}

\section{Preliminary Findings (Mock Data)}
Using a prototype evaluated against a generated PAN-style mock dataset, the following performance was observed:

\begin{table}[h!]
\centering
\begin{tabular}{@{}llll@{}}
\toprule
Metric & Baseline (Standard SBERT) & \textbf{Hybrid + Smith-Waterman} & Improvement \\ \midrule
Precision & 0.2677 & \textbf{0.4971} & +85\% \\
Recall & 1.0000 & \textbf{1.0000} & (Baseline) \\
F1 Score & 0.4223 & \textbf{0.6641} & +57\% \\ \bottomrule
\end{tabular}
\caption{Comparison of initial results after Threshold Tuning.}
\end{table}

\textbf{Analysis:} The current architecture maintains a perfect Recall (1.0), meaning it catches all plagiarism cases. The transition to the Smith-Waterman alignment significantly boosted Precision, reducing false positives by focusing on global document sequences rather than local window matches.

\section{Next Steps}
\begin{itemize}
    \item \textbf{Data Ingestion:} Transition from mock data to the full official \textbf{PAN 2025 Dataset}.
    \item \textbf{Weight Optimization:} Fine-tuning the balance between Semantic (70\%) and Structural (30\%) components.
    \item \textbf{Baseline Validation:} Running comparative tests against SVM and N-gram models as promised in Section 1.5 of the proposal.
\end{itemize}

\newpage
\section{FAQ: Supervisor Meeting Preparation}

\textbf{Q1: Why are we using a Hybrid approach instead of just a Transformer (BERT)?} \\
\textit{Answer:} Scientific articles have very rigid structures and specialized jargon. Transformers are great at meaning, but they can be "tricked" by synonym replacement in jargon. Adding structural features (like POS tag ratios) provides a stylistic "fingerprint" that makes the detection much more robust against sophisticated paraphrasing.

\textbf{Q2: Why is the Smith-Waterman algorithm necessary?} \\
\textit{Answer:} Standard similarity tools compare windows in isolation. Smith-Waterman is a dynamic programming algorithm that treats the document as a sequence. It allows us to detect the "optimal path" of similarity, effectively handling cases where a plagiarist has scrambled the order of paragraphs or inserted their own sentences in between (patchwriting).

\textbf{Q3: Is this architecture considered State-of-the-Art (SOTA)?} \\
\textit{Answer:} Yes. As identified in the literature review (e.g., Antonius et al., 2023), the combination of deep semantic embeddings with sequence alignment is the current benchmark for 99\% accuracy in document similarity.

\textbf{Q4: Why is the Precision currently only 0.49 on mock data?} \\
\textit{Answer:} The current mock dataset is very small, which often results in higher false-positive rates. However, the architecture is designed for large-scale data. Once the official PAN dataset is applied, we will perform hyperparameter optimization to reach the final SOTA scores.

\end{document}
